---
title: "Hypertension Analysis"
format: html
editor: visual
---

**DS400: Bayesian Statistics \| Ashley Sofia Alfaro**

**RQ:** What are the strongest indicators of hypertension (high blood pressure), and how accurately can we predict the probability that someone with similar habits would have hypertension?

**Dataset:** [Diabetes, Hypertension, and Stroke Prediction](https://www.kaggle.com/datasets/prosperchuks/health-dataset) on Kaggle

------------------------------------------------------------------------

## Libraries

```{r, message=FALSE}
library(janitor)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(here)
library(bayesrules)
library(bayesplot)
library(rstanarm)
library(shiny)
library(shinythemes)
```

## Read In Data

```{r}
hpt_df <- read.csv(here("data/diabetes_data.csv"))
```

## Mutate Data

This is to convert variables to factors with clear labels.

```{r}
hpt_df <- hpt_df %>%
  mutate(HighBP = factor(HighBP, labels = c("No Hypertension", "Hypertension"))) %>% 
  mutate(Diabetes = factor(Diabetes, labels = c("No Diabetes", "Diabetes"))) %>% 
  mutate(Age = factor(Age, labels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49",
                                      "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+"))) %>% 
  mutate(HighChol = factor(HighChol, labels = c("No", "Yes"))) %>% 
  mutate(Sex = factor(Sex, labels = c("Male", "Female"))) %>% 
  mutate(HeartDiseaseorAttack = factor(HeartDiseaseorAttack, labels = c("No", "Yes"))) %>%
  mutate(PhysActivity = factor(PhysActivity, labels = c("No", "Yes"))) %>% 
  mutate(Smoker = factor(Smoker, labels = c("No", "Yes"))) %>% 
  mutate(Fruits = factor(Fruits, labels = c("No", "Yes"))) %>% 
  mutate(Veggies = factor(Veggies, labels = c("No", "Yes")))
```

------------------------------------------------------------------------

## Exploratory Analysis

**Checking structure of the dataset**

```{r}
str(hpt_df)
```

**Looking at the unique BMI values to see its range**

```{r}
unique(hpt_df$BMI)
```

**Bar plot showing count of hypertension cases vs no hypertension**

```{r}
ggplot(hpt_df, aes(x = HighBP, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  count(HighBP)
```

**Bar plot of age groups filled by hypertension status**

```{r}
ggplot(hpt_df, aes(x = Age, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, Age) %>% 
  adorn_percentages("col")
```

**Interpretation:** Those younger have lower probabilities of having hypertension, but the probability flips when you reach the 50-54 age range and older. This shows that older groups have a much higher risk of hypertension.

**Bar plot showing BMI distribution and hypertension status**

```{r}
ggplot(hpt_df, aes(x = BMI, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

**Bar plot of Diabetes status vs hypertension**

```{r}
ggplot(hpt_df, aes(x = Diabetes, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, Diabetes) %>% 
  adorn_percentages("row")
```

**Interpretation:** Individuals with hypertension are likely to have diabetes, too.

**The code chunks below produce similar plots and tables for High Cholesterol, Heart Disease or Attack, Sex, and Physical Activity to understand their relationships with hypertension.**

```{r}
ggplot(hpt_df, aes(x = HighChol, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, HighChol) %>% 
  adorn_percentages("row")
```

```{r}
ggplot(hpt_df, aes(x = HeartDiseaseorAttack, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, HeartDiseaseorAttack) %>% 
  adorn_percentages("row")
```

```{r}
ggplot(hpt_df, aes(x = Sex, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
ggplot(hpt_df, aes(x = PhysActivity, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, PhysActivity) %>% 
  adorn_percentages("col")
```

------------------------------------------------------------------------

## Model

I want to use *weakly informative* priors for the Bayesian logistic regression model because I don't want to force the model to believe anything specific, but instead to give it reasonable starting assumptions so it would not produce extreme or unrealistic estimates. Basically, I don't want the posterior probability to be heavily based on our prior belief.

-   **Target:** HighBP
-   **Possible Predictors:** Age, BMI, Diabetes, HighChol, HeartDiseaseorAttack, PhysActivity

```{r}
hpt_df$Age <- as.factor(hpt_df$Age)

multi_hpt_model <- stan_glm(
  HighBP ~ Age + Sex + BMI + HeartDiseaseorAttack + PhysActivity + HighChol + Diabetes,
  data = hpt_df,
  family = binomial,
  prior_intercept = normal(0, 2),
  prior = normal(0, 1),
  chains = 4, iter = 4000, seed = 123
)

multi_hpt_model
```

### Accuracy

```{r}
classification_summary(model = multi_hpt_model, data = hpt_df, cutoff = 0.4)
```

-   Sensitivity (true positives)
-   Specificity (true negatives)

**Model Accuracy:** \~72.98% **Sensitivity:** \~88.22% **Specificity:** \~53.31%

------------------------------------------------------------------------

### Predictions for a Single Person

Creating a "person" with certain characteristics and having it predict the probability this person would have hypertension.

```{r}
our_person <- data.frame(Age = "50-54", Sex = "Male", BMI = 29, HeartDiseaseorAttack = "Yes", PhysActivity = "No", HighChol = "Yes", Diabetes = "No Diabetes")
```

```{r}
predict(multi_hpt_model, newdata = our_person, type = "response")
```

### Predictions for Dataset

Adding predicted probability to the dataset

```{r}
hpt_filtered_df <- hpt_df %>% 
  select(Age, Sex, BMI, HeartDiseaseorAttack, PhysActivity, HighChol, Diabetes, HighBP) %>% 
  mutate(predicted_hypertension = predict(multi_hpt_model, newdata = ., type = "response"))
```

Plot predicted probability vs actual hypertension status

```{r, fig.width=10}
ggplot(hpt_filtered_df, aes(x = predicted_hypertension, y = HighBP)) +
  geom_jitter(size = 1, alpha = 0.1) +
  theme_minimal()
```

### Posterior Intervals

```{r}
exp(posterior_interval(multi_hpt_model, prob = 0.8))
```

```{r}
plot(multi_hpt_model)
```

**Increase probability of hypertension:** (Age) 35-39 → 80+, Sex (Female), Higher BMI, HeartDiseaseorAttack = Yes, HighChol = Yes, Diabetes = Yes

**Decrease probability of hypertension:** PhysActivity = Yes

**Uncertain / No Clear Effect:** (Age) 25-29, 30-34

------------------------------------------------------------------------

## Shiny App

This shiny app is meant to be used as a tool to predict hypertension risk based on selected characteristics. It also lets users simulate outcomes for 1000 similar people to see variability in predictions.

-   **Note:** This is **not** intended to be a diagnostic tool.

Save model for use in the Shiny app

```{r}
saveRDS(multi_hpt_model, "multi_hpt_model.rds")
```

Load model and build Shiny UI and Server

```{r}
model <- readRDS("multi_hpt_model.rds")

ui <- fluidPage(
  theme = shinytheme("simplex"),
  titlePanel(strong("Hypertension Probability Calculator")),

  tags$style(HTML("
    #simulation_text {
      font-size: 14px;
      padding-top: 6px;
    }
    .section-title {
      margin-top: 25px;
      font-weight: bold;
      font-size: 20px;
    }
  ")),

  sidebarLayout(
    sidebarPanel(
      h4("Select Characteristics"),
      selectInput("age", "Age Group:", 
                  choices = levels(hpt_df$Age)),
      selectInput("sex", "Sex:", 
                  choices = levels(hpt_df$Sex)),
      selectInput("active", "Physically Active:", 
                  choices = levels(hpt_df$PhysActivity)),
      sliderInput("bmi", "BMI:", 
                  min = 15, max = 55, value = 25),
      selectInput("heart", "Heart Disease or Attack:", 
                  choices = levels(hpt_df$HeartDiseaseorAttack)),
      selectInput("chol", "High Cholesterol:", 
                  choices = levels(hpt_df$HighChol)),
      selectInput("diabetes", "Diabetes:", 
                  choices = levels(hpt_df$Diabetes)),
      actionButton("go", "Predict", class = "btn-primary"),
      br(), br(),
      actionButton("simulate", "Simulate 1000 People", class = "btn-success")
    ),

    mainPanel(
      div(class="section-title", "Predicted Probability"),
      wellPanel(
        style = "background-color:#f7f7f7; padding: 14px; border-radius: 6px;",
        textOutput("result")
      ),
      plotOutput("posterior_plot"),

      div(class="section-title", "Simulation Results"),
      wellPanel(
        style = "background-color:#f7f7f7; padding: 14px; border-radius: 6px;",
        textOutput("simulation_text")
      ),
      plotOutput("simulation_plot"),
      br()
    )
  )
)

server <- function(input, output, session){
  # Create a dataframe for the selected characteristics 
  observeEvent(input$go, {
    new_person <- tibble(
      Age = factor(input$age, levels = levels(hpt_df$Age)),
      Sex = factor(input$sex, levels = levels(hpt_df$Sex)),
      PhysActivity = factor(input$active, levels = levels(hpt_df$PhysActivity)),
      BMI = as.numeric(input$bmi),
      HeartDiseaseorAttack = factor(input$heart, levels = levels(hpt_df$HeartDiseaseorAttack)),
      HighChol = factor(input$chol, levels = levels(hpt_df$HighChol)),
      Diabetes = factor(input$diabetes, levels = levels(hpt_df$Diabetes))
    )

    prob_draws <- posterior_linpred(model, newdata = new_person, transform = TRUE)
    prob <- mean(prob_draws)

    output$result <- renderText({
      paste0("Estimated probability of hypertension: ", round(prob * 100, 1), "%")
    })

    output$posterior_plot <- renderPlot({
      density_obj <- density(prob_draws)
      plot(density_obj,
           main = "Posterior Distribution of Predicted Probability",
           xlab = "Probability of Hypertension",
           ylab = "Density",
           lwd = 2, col = "red", xlim = c(0,1))
      polygon(density_obj, col = rgb(1, 0, 0, 0.3), border = NA)
      abline(v = prob, col = "blue", lwd = 2, lty = 2)
    })
  })

  observeEvent(input$simulate, {
    new_person <- tibble(
      Age = factor(input$age, levels = levels(hpt_df$Age)),
      Sex = factor(input$sex, levels = levels(hpt_df$Sex)),
      PhysActivity = factor(input$active, levels = levels(hpt_df$PhysActivity)),
      BMI = as.numeric(input$bmi),
      HeartDiseaseorAttack = factor(input$heart, levels = levels(hpt_df$HeartDiseaseorAttack)),
      HighChol = factor(input$chol, levels = levels(hpt_df$HighChol)),
      Diabetes = factor(input$diabetes, levels = levels(hpt_df$Diabetes))
    )

    prob_draws <- posterior_linpred(model, newdata = new_person, transform = TRUE)

    simulated_outcomes <- rbinom(1000, size = 1, prob = prob_draws)

    output$simulation_plot <- renderPlot({
      sim_density <- density(simulated_outcomes)
      plot(sim_density,
           main = "Distribution of Outcomes (Based on Selected Characteristics)",
           xlab = "Simulated Hypertension Outcome\n 0 = No Hypertension; 1 = Hypertension",
           ylab = "Density",
           lwd = 2, col = "darkgreen",
           xlim = c(0,1))
      polygon(sim_density, col = rgb(0,1,0,0.3), border = NA)
    })

    output$simulation_text <- renderText({
      paste0("Out of 1000 simulated people with these characteristics, ",
             sum(simulated_outcomes),
             " were predicted to have hypertension.")
    })
  })
}

shinyApp(ui = ui, server = server)
```
