---
title: "Hypertension Analysis"
format: html
editor: visual
---

**DS400: Bayesian Statistics \| Ashley Sofia Alfaro**

**Objective:** What are the strongest indicators of hypertension (high blood pressure), and how accurately can we predict the probability that someone with similar habits would have hypertension?

**Dataset:** [Diabetes, Hypertension, and Stroke Prediction](https://www.kaggle.com/datasets/prosperchuks/health-dataset) on Kaggle

## Libraries

```{r, message=FALSE}
library(janitor)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(here)
library(bayesrules)
library(bayesplot)
library(rstanarm)
library(shiny)
```

## Read In Data

```{r}
hpt_df <- read.csv(here("data/diabetes_data.csv"))
```

## Mutate Data

```{r}
hpt_df <- hpt_df %>%
  mutate(HighBP = factor(HighBP, labels = c("No Hypertension", "Hypertension"))) %>% 
  mutate(Diabetes = factor(Diabetes, labels = c("No Diabetes", "Diabetes"))) %>% 
  mutate(Age = factor(Age, labels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49",
                                      "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+"))) %>% 
  mutate(HighChol = factor(HighChol, labels = c("No", "Yes"))) %>% 
  mutate(Sex = factor(Sex, labels = c("Male", "Female"))) %>% 
  mutate(HeartDiseaseorAttack = factor(HeartDiseaseorAttack, labels = c("No", "Yes"))) %>%
  mutate(PhysActivity = factor(PhysActivity, labels = c("No", "Yes"))) %>% 
  mutate(Smoker = factor(Smoker, labels = c("No", "Yes"))) %>% 
  mutate(Fruits = factor(Fruits, labels = c("No", "Yes"))) %>% 
  mutate(Veggies = factor(Veggies, labels = c("No", "Yes")))
```

## Exploratory Analysis

```{r}
str(hpt_df)
```

```{r}
unique(hpt_df$BMI)
```

```{r}
ggplot(hpt_df, aes(x = HighBP, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  count(HighBP)
```

```{r}
ggplot(hpt_df, aes(x = Age, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, Age) %>% 
  adorn_percentages("col")
```

Those younger have lower probabilities of having hypertension, but the probability flips when you reach the 50-54 age range and older. This shows that older groups have a much higher risk of hypertension.

```{r}
ggplot(hpt_df, aes(x = BMI, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
ggplot(hpt_df, aes(x = Diabetes, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, Diabetes) %>% 
  adorn_percentages("row")
```

Individuals with hypertension are likely to have diabetes, too.

```{r}
ggplot(hpt_df, aes(x = HighChol, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, HighChol) %>% 
  adorn_percentages("row")
```

```{r}
ggplot(hpt_df, aes(x = HeartDiseaseorAttack, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, HeartDiseaseorAttack) %>% 
  adorn_percentages("row")
```

```{r}
ggplot(hpt_df, aes(x = Sex, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
ggplot(hpt_df, aes(x = PhysActivity, fill = HighBP)) +
  geom_bar() +
  theme_minimal()
```

```{r}
hpt_df %>% 
  tabyl(HighBP, PhysActivity) %>% 
  adorn_percentages("col")
```

**Target:** HighBP\
**Possible Predictors:** Age, BMI, Diabetes, HighChol, HeartDiseaseorAttack, PhysActivity

## Model

I want to use weakly informative priors for the Bayesian logistic regression model because I don't want to force the model to believe anything specific, but instead to give it reasonable starting assumptions so it would not produce extreme or unrealistic estimates. Basically, I don't want the posterior probability to be heavily based on our prior belief.

-   **normal(0, 2)** This works well when there is no strong prior belief about the baseline probability of the outcome. It keeps the model stable without restricting it too much.

-   **normal(0, 1)** This would help prevent overfitting or exaggerated coefficients, especially for factors with smaller sample sizes.

```{r}
hpt_df$Age <- as.factor(hpt_df$Age)

multi_hpt_model <- stan_glm(
  HighBP ~ Age + Sex + BMI + HeartDiseaseorAttack + PhysActivity + HighChol + Diabetes,
  data = hpt_df,
  family = binomial,
  prior_intercept = normal(0, 2),
  prior = normal(0, 1),
  chains = 4, iter = 4000, seed = 123
)

multi_hpt_model
```

### Accuracy

```{r}
classification_summary(model = multi_hpt_model, data = hpt_df, cutoff = 0.4)
```

-   sensitivity (true positives)
-   specificity (true negatives)

Creating a "person" with certain characteristics and having it predict the probability this person would have hypertension.

```{r}
our_person <- data.frame(Age = "50-54", Sex = "Male", BMI = 29, HeartDiseaseorAttack = "Yes", PhysActivity = "No", HighChol = "Yes", Diabetes = "No Diabetes")
```

```{r}
predict(multi_hpt_model, newdata = our_person, type = "response")
```

Using the model to predict the probability someone has hypertension within the dataset.

```{r}
hpt_filtered_df <- hpt_df %>% 
  select(Age, Sex, BMI, HeartDiseaseorAttack, PhysActivity, HighChol, Diabetes, HighBP) %>% 
  mutate(predicted_hypertension = predict(multi_hpt_model, newdata = ., type = "response"))
```

```{r, fig.width=10}
ggplot(hpt_filtered_df, aes(x = predicted_hypertension, y = HighBP)) +
  geom_jitter(size = 1, alpha = 0.1) +
  theme_minimal()
```

### Posterior Distributions

```{r}
posterior_interval(multi_hpt_model)
```

```{r}
plot(multi_hpt_model)
```

**Increase probability of hypertension:** (Age) 35-39 → 80+, Sex (Male), Higher BMI, HeartDiseaseorAttack = Yes, HighChol = Yes, Diabetes = Yes

**Decrease probability of hypertension:** PhysActivity = Yes

pp_check(multi_hpt_model, plotfun = "hist")

pp_check(multi_hpt_model, plotfun = "scatter_avg")

**Uncertain / No Clear Effect:** (Age) 25-29, 30-34
