---
title: "Hypertension Probability Calculator"
format: html
editor: visual
---

## Libraries
```{r}
library(shiny)
library(shinythemes)
library(here)
library(dplyr)
```

## Read and Mutate Data
```{r}
hpt_df <- read.csv(here("data/diabetes_data.csv"))
```

```{r}
hpt_df <- hpt_df %>%
  mutate(HighBP = factor(HighBP, labels = c("No Hypertension", "Hypertension"))) %>% 
  mutate(Diabetes = factor(Diabetes, labels = c("No Diabetes", "Diabetes"))) %>% 
  mutate(Age = factor(Age, labels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49",
                                      "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+"))) %>% 
  mutate(HighChol = factor(HighChol, labels = c("No", "Yes"))) %>% 
  mutate(Sex = factor(Sex, labels = c("Male", "Female"))) %>% 
  mutate(HeartDiseaseorAttack = factor(HeartDiseaseorAttack, labels = c("No", "Yes"))) %>%
  mutate(PhysActivity = factor(PhysActivity, labels = c("No", "Yes"))) %>% 
  mutate(Smoker = factor(Smoker, labels = c("No", "Yes"))) %>% 
  mutate(Fruits = factor(Fruits, labels = c("No", "Yes"))) %>% 
  mutate(Veggies = factor(Veggies, labels = c("No", "Yes")))
```


## Shiny App
This shiny app is meant to be used as a tool to predict hypertension risk based on selected characteristics. It also lets users simulate outcomes for 1000 similar people to see variability in predictions.

-   **Note:** This is **not** intended to be a diagnostic tool.

Make sure you have already downloaded the ``hpt_model.rds`` file, then run the code chunk below to get and interact with the shiny app!

```{r}
model <- readRDS("hpt_model.rds")

ui <- fluidPage(
  theme = shinytheme("simplex"),
  titlePanel(strong("Hypertension Probability Calculator")),

  tags$style(HTML("
    #simulation_text {
      font-size: 14px;
      padding-top: 6px;
    }
    .section-title {
      margin-top: 25px;
      font-weight: bold;
      font-size: 20px;
    }
  ")),

  sidebarLayout(
    sidebarPanel(
      h4("Select Characteristics"),
      selectInput("age", "Age Group:", 
                  choices = levels(hpt_df$Age)),
      selectInput("sex", "Sex:", 
                  choices = levels(hpt_df$Sex)),
      selectInput("active", "Physically Active:", 
                  choices = levels(hpt_df$PhysActivity)),
      sliderInput("bmi", "BMI:", 
                  min = 15, max = 55, value = 25),
      selectInput("heart", "Heart Disease or Attack:", 
                  choices = levels(hpt_df$HeartDiseaseorAttack)),
      selectInput("chol", "High Cholesterol:", 
                  choices = levels(hpt_df$HighChol)),
      selectInput("diabetes", "Diabetes:", 
                  choices = levels(hpt_df$Diabetes)),
      actionButton("go", "Predict", class = "btn-primary"),
      br(), br(),
      actionButton("simulate", "Simulate 1000 People", class = "btn-success")
    ),

    mainPanel(
      div(class="section-title", "Predicted Probability"),
      wellPanel(
        style = "background-color:#f7f7f7; padding: 14px; border-radius: 6px;",
        textOutput("result")
      ),
      plotOutput("posterior_plot"),

      div(class="section-title", "Simulation Results"),
      wellPanel(
        style = "background-color:#f7f7f7; padding: 14px; border-radius: 6px;",
        textOutput("simulation_text")
      ),
      plotOutput("simulation_plot"),
      br()
    )
  )
)

server <- function(input, output, session){
  # Create a dataframe for the selected characteristics 
  observeEvent(input$go, {
    new_person <- tibble(
      Age = factor(input$age, levels = levels(hpt_df$Age)),
      Sex = factor(input$sex, levels = levels(hpt_df$Sex)),
      PhysActivity = factor(input$active, levels = levels(hpt_df$PhysActivity)),
      BMI = as.numeric(input$bmi),
      HeartDiseaseorAttack = factor(input$heart, levels = levels(hpt_df$HeartDiseaseorAttack)),
      HighChol = factor(input$chol, levels = levels(hpt_df$HighChol)),
      Diabetes = factor(input$diabetes, levels = levels(hpt_df$Diabetes))
    )

    prob_draws <- posterior_linpred(model, newdata = new_person, transform = TRUE)
    prob <- mean(prob_draws)

    output$result <- renderText({
      paste0("Estimated probability of hypertension: ", round(prob * 100, 1), "%")
    })

    output$posterior_plot <- renderPlot({
      density_obj <- density(prob_draws)
      plot(density_obj,
           main = "Posterior Distribution of Predicted Probability",
           xlab = "Probability of Hypertension",
           ylab = "Density",
           lwd = 2, col = "red", xlim = c(0,1))
      polygon(density_obj, col = rgb(1, 0, 0, 0.3), border = NA)
      abline(v = prob, col = "blue", lwd = 2, lty = 2)
    })
  })

  observeEvent(input$simulate, {
    new_person <- tibble(
      Age = factor(input$age, levels = levels(hpt_df$Age)),
      Sex = factor(input$sex, levels = levels(hpt_df$Sex)),
      PhysActivity = factor(input$active, levels = levels(hpt_df$PhysActivity)),
      BMI = as.numeric(input$bmi),
      HeartDiseaseorAttack = factor(input$heart, levels = levels(hpt_df$HeartDiseaseorAttack)),
      HighChol = factor(input$chol, levels = levels(hpt_df$HighChol)),
      Diabetes = factor(input$diabetes, levels = levels(hpt_df$Diabetes))
    )

    prob_draws <- posterior_linpred(model, newdata = new_person, transform = TRUE)

    simulated_outcomes <- rbinom(1000, size = 1, prob = prob_draws)

    output$simulation_plot <- renderPlot({
      sim_density <- density(simulated_outcomes)
      plot(sim_density,
           main = "Distribution of Outcomes (Based on Selected Characteristics)",
           xlab = "Simulated Hypertension Outcome\n 0 = No Hypertension; 1 = Hypertension",
           ylab = "Density",
           lwd = 2, col = "darkgreen",
           xlim = c(0,1))
      polygon(sim_density, col = rgb(0,1,0,0.3), border = NA)
    })

    output$simulation_text <- renderText({
      paste0("Out of 1000 simulated people with these characteristics, ",
             sum(simulated_outcomes),
             " were predicted to have hypertension.")
    })
  })
}

shinyApp(ui = ui, server = server)
```